\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-18 docs/parco13/abstract.tex
Go through parco13 abstract

\newcommand{\comment}[1]{\textcolor{green}{[DAC: #1]}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\clawpack}{{\sc Clawpack\xspace}}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, finite volume (element?)
\cite{???}, or discontinuous Galerkin
\cite{BursteddeGhattasGurnisEtAl10} methods.  The finite volume method
has typically been implemented using a single degree of freedom per
cell on either structured \cite{ppm, clawpack, weno} or unstructured
meshes \cite{ddfv, ausm+, openfoam, }.  \comment{Are you looking for
  references to parallel AMR codes?  Or more general AMR codes based
  on the FVM?}  \todo{No, I was looking for unstructured FV
  references.  Would you have any, and can we add the block-AMR
  references later?} \comment{But most finite volume methods
  (structured or unstructured) only have a single degree of freedom
  per mesh cell (``element''?)  Only recently have people increased
  the order by extending the stencil (without increase the DOFs per
  mesh cell).}

@@

To enable hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989, be-le:1991,
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping (except in ghost cell regions),
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% Clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC : Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf; \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I hav added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed.
\end{enumerate}

% yields periodicity as a special case
A particular feature of \forestclaw is that the generic handling of multiblock
geometries is inherited from \pforest, identifying each octree as a block, and
each leaf as a patch.  Each block is understood as a reference unit cube with
its own geometric mapping.  The connectivity of the blocks is static and can be
created by external hexahedral mesh generators, eliminating the need to encode
it by hand.

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of ghost
cells surrounding each patch.  To this end, we define an interface
that allows read-only access to the sequence of blocks,
the list of patches for each, and a lookup of neighbor patches (and
their relative orientation which is nontrivial between blocks).
Just as \pforest, \forestclaw stores only the patches local to each
MPI rank.  The exchange of ghost data and parallel partitioning is provided
transparently by \pforest.  Mesh modification directives, such as
local refinement and coarsening, are called from \forestclaw and
relayed to \pforest.

% Forestclaw would support swapping out p4est or Clawpack

@@

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The order of leaves is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost patches.

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
subcycling).  When using subcycling, the load balance is attained per level as
it is done in structured AMR or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch is handled by
\forestclaw alone without the need to involve \pforest.

@@

For hyperbolic problems, we integrate the solution on a a single patch
using the wave propagation algorithm described by R. J. LeVeque and
implemented in {\sc Clawpack} \cite{le:2002, clawpack}.  We assume a
single degree of freedom per mesh cell and reconstruct a piecewise
constant solution to obtain left and right states at a mesh cell
edges.  At each edge, we solve Riemann problems to obtain left and
right going waves propagating at speeds determined from the solution
to the Riemann problem.  For scalar advection, the speed of each wave
is simply the local advection speed at the cell interface. For
non-linear problems and systems, an approximate Riemann solver, such
as a Roe solver, is typically used. Wave limiters are used to surpress
spurious oscillations at sharp gradients in the flow.

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When subcycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.

Grid refinement and de-refinement also requires interpolation from coarse grids
to newly created fine grids, and the averaging of fine grid data to a newly created
underlying coarse grid.

@@

The authors acknowledge valuable discussion with Randy LeVeque, Marsha Berger,
and Hans-Petter Langtangen.  We also acknowledge David Ketchison and the KAUST sponsored HPC$^3$
numerics workshop at which the initial phases of this project were first discussed.
The leaf/patch paradigm was independently presented
by B.\ as part of a talk at the SCI Institute, Utah, in July 2011.

>>>>

\newcommand{\comment}[1]{\textcolor{green}{[DAC: #1]}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\clawpack}{{\sc Clawpack}\xspace}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}
\newcommand{\manyclaw}{Manyclaw\xspace}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack, weno}
or unstructured meshes
\cite{ddfv, ausm+, openfoam},
where higher order methods have been constructed by widening the stencil
\cite{needonehere}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

% yields periodicity as a special case
A particular feature of \forestclaw is that the generic handling of multiblock
geometries is inherited from \pforest, identifying each octree as a block.
% and each leaf as a patch.
Each block is understood as a reference unit cube with
its own geometric mapping.  The connectivity of the blocks can be created by
external hexahedral mesh generators, eliminating the need to encode it by hand.

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of ghost
cells surrounding each patch.  To this end, we define an interface
that allows read-only access to the sequence of blocks,
the list of patches for each, and a lookup of neighbor patches (and
their relative orientation which is nontrivial between blocks).
Suitably informed by \pforest, \forestclaw stores only the patches local to
each MPI rank.  The exchange of ghost data and parallel repartitioning is
provided transparently by \pforest.  Mesh modification directives, such as
adaptive refinement and coarsening, are called from \forestclaw and relayed to
\pforest.

% Forestclaw would support swapping out p4est or \clawpack

@@

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The order of leaves is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost leaves.

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
subcycling).  When using subcycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.

@@

% CB: I'm trying to avoid "mesh" cell here since in \pforest FE simulations
% that would mean a whole leaf.

For hyperbolic problems, we integrate the solution on a a single uniform
patch, containing $m^d$ cells,
using the wave propagation algorithm described by R. J. LeVeque and
implemented in \clawpack \cite{le:2002, clawpack}.  We assume a
single degree of freedom per cell and reconstruct a piecewise
constant solution to obtain left and right states at cell
edges.  At each edge, we solve Riemann problems to obtain left and
right going waves propagating at speeds determined from the solution
to the Riemann problem.  For scalar advection, the speed of each wave
is simply the local advection speed at the cell interface.  For non-linear
problems and systems, an approximate Riemann solver, such as a Roe solver
\cite{roesolver}, is typically used.  Wave limiters are used to surpress
spurious oscillations at sharp gradients in the flow.

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  This implies that
one layer of ghost patches is sufficient for $m \ge 4$.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When subcycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.
As mentioned in the previous section, this procedure can be extended
transparently to distributed parallelism by defining an abstract exchange
routine for ghost patch data.

Grid refinement and coarsening requires interpolation from coarse grids
to newly created fine grids, and the averaging of fine grid data to a newly
created underlying coarse grid.  This operation is rank-local, analogous to the
general dynamic AMR procedures used in \pforest-based FE codes \cite[Fig.\
4]{BursteddeGhattasStadlerEtAl08}.

@@

The authors acknowledge valuable discussion with Randy LeVeque, Marsha Berger,
and Hans-Petter Langtangen.  We also acknowledge David Ketcheson and the KAUST
sponsored HPC$^3$ numerics workshop at which the initial phases of this project
were first discussed.
The leaf/patch paradigm was independently presented
by B.\ as part of a talk at the SCI Institute, Utah, in July 2011.
